A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-31 latex/main.tex
Updates from Overleaf

The current HMM assumes a simplistic mutation model
(where mutation to any other allele is equally probable) and
implicitly assumes a uniform mutation rate along the SARS-CoV-2 genome.
However, genome analyses have revealed
transition-to-transversion rate bias in SARS-CoV-2 \cite{Bloom2023} and
sequence context-dependent variation
in the mutation rate along the SARS-CoV-2 genome (REF).
A more realistic mutation model may help improve
inference of local genetic relationships in the ARG, and
accounting for mutation rate heterogeneity
during sample matching or in the HMM cost
(e.g., down-weighing sites in highly mutable regions)
may help reduce false positive recombinants.

We developed and used several methods
to exclude poor-quality recombination events post hoc, including
filtering by supporting loci per parent and
long-branch splitting.
These methods could instead be run immediately
after sample matching during ARG inference,
obviating the need for post-processing of the ARG.

An inherent limitation of detecting recombination from sequence data,
which applies to all methods,
is that it is impossible to detect events with few (or no) supporting bases,
as these cannot be distinguished from mutation events.
This limitation makes it particularly challenging
to detect events near the termini of linear viral genomes.

It is challenging to accurately reconstruct global events
when there are regions with poor sampling and jurisdictions
with long delays before sequences become publicly available (REF).
Revisiting the ARG topology periodically
(e.g., rewinding the ARG a month and repeating
with samples that became available later) and
better handling of putative recombinants
with suspiciously long branches (e.g., Delta and Omicron BA.2)
might improve the accuracy of the genealogy.

@@

@@

SHZ is supported by an Intermediate Research Fellowship
from the Nuffield Department of Population Health,
University of Oxford.

@@

S.Y.W. is an employee and shareholder of COMPANY.

>>>>

The current HMM assumes a simplistic mutation model
(where mutation to any other allele is equally probable) and
implicitly assumes a uniform mutation rate along the SARS-CoV-2 genome.
However, genome analyses have revealed
transition-to-transversion rate bias \cite{Bloom2023,Simmonds2024} and
sequence context-dependent variation
in the mutation rate in the SARS-CoV-2 genome \cite{Haddox2025}.
A more realistic mutation model may help improve
inference of local genetic relationships in the ARG, and
accounting for mutation rate heterogeneity
during sample matching or in the HMM cost
(e.g., down-weighing sites in highly mutable regions)
may help reduce false positive recombinants.

@@

It is challenging to accurately reconstruct global events
when there are regions with poor sampling and
jurisdictions with long delays
before sequences become publicly available (REF).
For example, revisiting the ARG topology periodically
(e.g., rewinding the ARG a month and repeating
with samples that became available later)
might improve the accuracy of the genealogy.

Several methods are developed and used
to exclude probable false positive recombination events \textit{post hoc},
including filtering by supporting loci per parent and
splitting of suspiciously long branches.
These methods could instead be run immediately
after sample matching during ARG inference,
obviating the need for post-processing of the ARG.

Finally, an inherent limitation of detecting recombination from sequence data,
which applies to all methods,
is that it is impossible to detect events with few (or no) supporting bases,
as these cannot be distinguished from mutation events.
This limitation makes it particularly challenging
to detect events near the termini of linear viral genomes.

@@

SHZ is supported by an NDPH Intermediate Research Fellowship
from Oxford Population Health, University of Oxford.

@@

% TODO: Close https://github.com/jeromekelleher/sc2ts-paper/issues/922
The authors declare no competing interests.
